---
title: 'Version 0.2'
output: 
  flexdashboard::flex_dashboard:
    source_code: https://github.com/scott-saunders/growth_curve_fitting/blob/master/growth_curve_fitting_ver0.2.Rmd
fontsize: 12pt
runtime: shiny
---

```{r setup}
library(tidyverse)
library(dplyr)
library(cowplot)
library(viridis)
library(knitr)
library(kableExtra)
library(modelr)
library(broom)
library(ggforce)
library(grofit)

```


# Column {.sidebar data-width='300%'}

## Growth Curve Fitting

Use this app to fit growth curves, whether it is one experiment or hundreds (e.g. from a plate reader).

-------

### Instructions

Start with some [example data](https://github.com/scott-saunders/labwork/blob/master/Growth_curve_fitting/data.zip).

1. **Upload** a .csv file of your data 
2. **Select Data** to be fit.
3. **Set Model Parameters** with a reasonable guess.
4. Examine **Model Fits** and check for errors.
5. View **Parameter Estimates**.
  * Use parameter estimates to inform model parameters for error datasets.
6. **Download** parameter estimates / best fit curves.




Enjoy! [-Scott Saunders](https://scott-saunders.github.io/website/)


-------

### Upload your data

Upload a .csv file where column 1 is "time" (in hrs). Data columns for each well / experiment follow. 

```{r}
fileInput("file1", label = '', multiple = FALSE, 
          accept = c("text/csv", "text/comma-separated-values,text/plain",".csv"))

df <- reactive({
  
  req(input$file1)
  
  df <- read_csv(input$file1$datapath)
  
  df_well_index <- tibble(well = colnames(df)[-1]) %>% mutate(well_index = row_number())

  df_wIndex <- left_join(df %>% gather(key = 'well',value = 'absorbance',-time), 
                         df_well_index, 
                         by = 'well' )
  
  
  return(df_wIndex)
  
})

```

-------

### Downloads

----

Parameter estimates of A, Lag, and Mu. Confidence intervals are included for gompertz models.

```{r}
 downloadHandler(
         filename = 'growth_curve_parameter_estimates.csv',
         outputArgs = list(label = 'Parameter estimates'),
         content = function(file) {
             write_csv(df_ests() %>% select(-c(data, fit_result, fit_errors, models, est_result, est_errors)), file)
         }
     )

```

-----

Best fit curves along with raw data.

```{r}

pred_file <- reactive({  
  
  pred_file <- df_preds()

  # if(input$biGomp){
  # 
  #   pred_file <- df_pred_ests()
  #   
  #   }
  
  })


downloadHandler(
  filename = 'growth_curve_model_fits.csv',
  outputArgs = list(label = 'Best fit curves'),
  content = function(file) { write_csv(pred_file() %>% select(-c(data, fit_result, fit_errors, models)), file)}
    )



```


# Select Data{ data-orientation=rows}

## Plots {data-height='300%' .tabset}

### Plot All


```{r}

output$overview_plot <- renderPlot({
  
  ggplot(df(), aes(time, absorbance, group = well)) + geom_path() + labs(title = 'Drag the box to select the data you would like to fit.')
  
})

plotOutput("overview_plot", brush = brushOpts(id = "overview_plot_brush"))

# Reactive that returns the whole dataset if there is no brush

selectedData <- reactive({
  data <- brushedPoints(df(), input$overview_plot_brush) 
  if (nrow(data) == 0)
    data <- df()
  data %>% filter(well %in% input$selected_wells)
})

```

### Plot Selected

```{r}

#output$select_plot <- 
  
  renderPlot({
  
  ggplot(selectedData(), aes(time, absorbance, group = well)) + geom_path()
  
})

#plotOutput("select_plot", height = '300px')

#renderUI(list(tags$hr(),plotOutput("overview_plot", brush = brushOpts(id = "overview_plot_brush"), height = '50px'), tags$hr(), plotOutput("select_plot", height = '100px')))

```

## Select Wells{data-height='100%'}

### Select Wells / Experiments

```{r}
renderUI({
  list(
    sliderInput(inputId = 'well_range', label = 'Select a range of wells', min = 1, max = max(df()$well_index), value = c(1,max(df()$well_index)))
  )
})

```

### Check boxes

```{r}
renderUI({    checkboxGroupInput(inputId = 'selected_wells', label = 'Choose your curves',
                     choices = levels(factor((df() %>% filter(well_index >= input$well_range[1]))$well)),
                     selected = levels(factor((df() %>% filter(well_index >= input$well_range[1]))$well)),
                     width = '100%', inline = T)})

```

## Tables {data-height='100%' .tabset}

### Table All


```{r}
renderTable({
  
  df() %>% pivot_wider(id_cols = 'time', names_from = 'well', values_from = 'absorbance' )
  
})
```

### Table Selected


```{r}
renderTable({
  selectedData() %>% pivot_wider(id_cols = 'time', names_from = 'well', values_from = 'absorbance' )
})
```


# Set Model Parameters{data-orientation=rows}

## Background 

### Background on Gompertz model parameters

```{r}
selectInput("model", "Select a model:", c("Single Gompertz" = "gomp", "Bi-Gompertz" = "biGomp","Spline" = "spline"), selected = 'gomp')
```

[This paper](https://aem.asm.org/content/56/6/1875), it suggests that the modified Gompertz equation is an ideal way to parameterize a growth curve. They desribe the three parameters as follows: 

>The three phases of the growth curve can be described by three parameters: the maximum specific growth rate, $\mu_m$ is defined as the tangent in the inflection point; the lag time, $Lag$, is defined as the x-axis intercept of this tangent; and the asymptote $A$ is the maximal value reached.

The equation used here is the following:

$$ y = A*exp \left(-exp \left(\frac{\mu_m}{A}(Lag - t) + 1 \right)\right) + C$$

Where $C$ is a starting absorbance value. 

The models are actually fit with a nonlinear least squares method, specifically the [nls function](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/nls) from R. This method is extremely fast, but it requires pretty good guesses for parameter values. Therefore you can provide custom parameter values below. If many of your experiments cannot be fit, try playing with these starting parameters.

## Gompertz model parameters {.tabset}

### Start parameters {style="height:50pc;"}

#### You have selected this model:

```{r}
renderText({
  case_when(
    input$model == "gomp" ~ "Single Gompertz",
    input$model == "biGomp" ~ "Bi-Gompertz",
    input$model == "spline" ~ "Spline" 
  )
})
```

#### Set starting parameter values

```{r}

params_gomp <- list(
      numericInput("A", "A start value", 0.5), 
      numericInput("Lag", "Lag start value", 4),
      numericInput("Mu", "Mu start value", 0.2), 
      numericInput("C", "C start value", 0.05),
      NULL,
      NULL,
      NULL
    )

params_biGomp <- list(
      numericInput("A", "A_1 start value", 0.5), 
      numericInput("A_2", "A_2 start value", 0.2),
      numericInput("Lag", "Lag_1 start value", 4),
      numericInput("Lag_2", "Lag_2 start value", 5),
      numericInput("Mu", "Mu_1 start value", 0.2), 
      numericInput("Mu_2", "Mu_2 start value", 0.3),
      numericInput("C", "C start value", 0.05)
    )

params_spline <- list(renderText("No params needed"),NULL, NULL, NULL, NULL, NULL, NULL )

renderUI({ 
  case_when(
    input$model == 'gomp' ~ params_gomp,
    input$model == 'biGomp' ~ params_biGomp,
    input$model == 'spline' ~ params_spline
  )
})
```


### Parameter Bounds {style="height:100pc;"}

#### You have selected this model:

```{r}
renderText({
  case_when(
    input$model == "gomp" ~ "Single Gompertz",
    input$model == "biGomp" ~ "Bi-Gompertz",
    input$model == "spline" ~ "Spline" 
  )
})
```


#### Set parameter bounds

```{r}
bounds_gomp <- list(
  numericInput("A_low", "A Lower bound", 0),
  numericInput("A_up", "A Upper bound", 100),
  numericInput("Lag_low", "Lag Lower bound", 0),
  numericInput("Lag_up", "Lag Upper bound", 100),
  numericInput("Mu_low", "Mu Lower bound", 0),
  numericInput("Mu_up", "Mu Upper bound", 10),
  numericInput("C_low", "C Lower bound", -10),
  numericInput("C_up", "C Upper bound", 10),
  NULL,
  NULL,
  NULL,
  NULL,
  NULL,
  NULL
)

bounds_biGomp <- list(
  numericInput("A_low", "A_1 Lower bound", 0),
  numericInput("A_up", "A_1 Upper bound", 100),
  numericInput("A_2_low", "A_2 Lower Bound", 0),
  numericInput("A_2_up", "A_2 Upper Bound", 100),
  numericInput("Lag_low", "Lag_1 Lower bound", 0),
  numericInput("Lag_up", "Lag_1 Upper bound", 100),
  numericInput("Lag_2_low", "Lag_2 Lower Bound", 0),
  numericInput("Lag_2_up", "Lag_2 Upper Bound", 100),
  numericInput("Mu_low", "Mu_1 Lower bound", 0),
  numericInput("Mu_up", "Mu_1 Upper bound", 10),
  numericInput("Mu_2_low", "Mu_2 Lower Bound", 0),
  numericInput("Mu_2_up", "Mu_2 Upper Bound", 10),
  numericInput("C_low", "C Lower bound", -10),
  numericInput("C_up", "C Upper bound", 10)
)

bounds_spline <- list( 
  renderText("No params needed"),
  NULL,
  NULL,
  NULL,
  NULL,
  NULL,
  NULL,
  NULL,
  NULL,
  NULL,
  NULL,
  NULL,
  NULL,
  NULL
)

renderUI({ 
  case_when(
    input$model == 'gomp' ~ bounds_gomp,
    input$model == 'biGomp' ~ bounds_biGomp,
    input$model == 'spline' ~ bounds_spline
  )
})
```


# Model Fits{data-orientation=rows}

```{r}
# This is where the models are actually fit

df_models <- reactive({
        
  
        # group and nest the data
        df_nest <- selectedData() %>%
            group_by(well) %>% 
            nest()
        
        
        # The nls function used to fit the data

        
        fit_gompertz_mod <- function(df){
            
            #A = 1; mu = 0.1; lam = 2
            
            mod <- nls(formula=absorbance~A*exp(-exp((Mu/A)*(Lag - time) + 1))+C, 
                       start = c(A = input$A, Mu = input$Mu, Lag = input$Lag, C = input$C), 
                       algorithm = 'port',
                       lower = c(A = input$A_low, Mu = input$Mu_low, Lag = input$Lag_low, C = input$C_low),
                       upper = c(A = input$A_up, Mu = input$Mu_up, Lag = input$Lag_up, C = input$C_up),
                       data = df)
            
            mod
        }
        
        fit_biGompertz_mod <- function(df){
            
            mod <- nls( formula=absorbance~( A_1*exp(-exp((Mu_1/A_1)*(Lag_1 - time) + 1)) ) + 
                                           ( A_2*exp(-exp((Mu_2/A_2)*(Lag_2 - time) + 1)) ) + C,
                        
#                        start = c(A_1 = 0.5, Mu_1 = 0.2, Lag_1 = 4, C= 0.05, 
#                                  A_2 = 0.2, Mu_2 = 0.3, Lag_2 = 5),
                        start = c(A_1 = input$A, Mu_1 = input$Mu, Lag_1 = input$Lag, C = input$C, 
                                  A_2 = input$A_2, Mu_2 = input$Mu_2, Lag_2 = input$Lag_2),
                        lower = c(A_1 = input$A_low, A_2 = input$A_2_low, Mu_1 = input$Mu_low, Mu_2 = input$Mu_2_low, Lag_1 = input$Lag_low, Lag_2 = input$Lag_2_low, C = input$C_low),
                        upper = c(A_1 = input$A_up, A_2 = input$A_2_up, Mu_1 = input$Mu_up, Mu_2 = input$Mu_2_up, Lag_1 = input$Lag_up, Lag_2 = input$Lag_2_up, C = input$C_up),
                        data = df, 
                        
                        algorithm = 'port')
            
            mod
          
          }

fit_spline_mod <- function(df){

  mod <- gcFitSpline(time = df$time, data = df$absorbance, gcID = 'best spline', control = grofit.control(interactive = F))

  mod

  }

  df_mod<- if(input$model == 'gomp'){
    
    df_nest %>% mutate(fit_result = map(data, safely(fit_gompertz_mod)))
    
  }else{
    if(input$model == 'biGomp'){
      
        df_nest %>% mutate(fit_result = map(data, safely(fit_biGompertz_mod)))

    } else{
        
      df_nest %>% mutate(fit_result = map(data, safely(fit_spline_mod)))

    }
  }

        # fit data using the above function ...do this safely in case of errors
       df_models <- df_mod %>% 
            ungroup() %>%
            mutate(fit_errors = map(fit_result, 'error')) %>%
             mutate(models = map(fit_result,'result')) %>%
             mutate(fit_error_occured = map(.x = models, .f = is.null), fit_error_occured = unlist(fit_error_occured))
        
        return(df_models)
        #return(df_mod)
        
    })
```


```{r}
    ##### generate the predictions from the models
    df_preds <- reactive({
        
        df_preds <- if(input$model == 'gomp' | input$model == 'biGomp'){
          
          df_models() %>% 
            filter(fit_error_occured == FALSE) %>% 
            mutate(preds = map2(data, models, add_predictions)) %>% 
            unnest(preds)
          
        } else{
          
          tidy_best_spline_preds <- function(df){
            
            tibble(time = df$fit.time,
                   pred = df$fit.data,
                   absorbance = df$raw.data)
            }
          
          df_models() %>% 
            filter(fit_error_occured == FALSE) %>% 
            mutate(preds = map(models, tidy_best_spline_preds)) %>% 
            unnest(preds)
        }
            
        return(df_preds)
    })
```

```{r}

    ###### generate the parameter estimates from the models
    df_ests <- reactive({
        df_ests <- if(input$model == 'gomp' | input$model == 'biGomp'){
          
          df_models() %>% 
            filter(fit_error_occured == FALSE) %>% 
            mutate(est_result = map(models, safely(tidy), conf.int = T)) %>% 
            mutate(est_errors = map(est_result, 'error')) %>% 
            mutate(ests = map(est_result,'result')) %>% 
            mutate(est_error_occured = map(.x = ests, .f = is.null), est_error_occured = unlist(est_error_occured)) %>% 
            mutate(ests = ifelse(est_error_occured == T, map(models, tidy, conf.int = F), ests)) %>% 
            unnest(ests)
        } else{
          
                    tidy_best_spline_ests <- function(df){
                      tibble(Lag = df$parameters$lambda, Mu = df$parameters$mu, A = df$parameters$A) %>% 
                        pivot_longer(cols = c(Lag, Mu, A),names_to = 'term',values_to = 'estimate') %>% 
                        filter(!is.na(estimate))
                    }
                    
                    df_models() %>% 
                      filter(fit_error_occured == FALSE) %>% 
                      mutate(ests = map(models, tidy_best_spline_ests )) %>% 
                      mutate(est_error_occured = T) %>% 
                      unnest(ests)

        }
        return(df_ests)
        
    })

```

```{r}
df_pred_ests <- reactive({

  df_pred_ests <- df_preds()

  if(input$model == 'biGomp'){

    gen_gompertz <- function(a, mu, lag, c, time){

      a*exp(-exp((mu/a)*(lag - time) + 1 )) + c

      }

    df_pred_ests <- left_join(df_preds(),
                              df_ests() %>%
                                pivot_wider(id_cols = 'well',
                                            names_from = 'term',
                                            values_from = 'estimate' ),
                              by = 'well') %>%
      mutate(pred_1 = gen_gompertz(a = A_1, mu = Mu_1, lag = Lag_1, c = C, time = time)) %>%
      mutate(pred_2 = gen_gompertz(a = A_2, mu = Mu_2, lag = Lag_2, c = C, time = time))

  }
  


    return(df_pred_ests)
})

#renderDataTable(df_pred_ests() %>% select(well, fit_error_occured, A_2))

```



## Plot overviews{.tabset}

### Fit Plots
```{r}

plotOutput("plot")

output$plot <- renderPlot({
  
  p <- ggplot(data = df_preds(), aes_string(x=df_preds()$time, y = df_preds()$absorbance)) + 
    geom_point(shape = 21) + 
    geom_path(aes(y = df_preds()$pred), size = 1, color = 'blue')+ 
    facet_wrap(~well, scales = 'free') +
    labs(x = 'Time (Hrs)', y = 'Absorbance') + theme_bw()
  
  if(input$model == 'biGomp'){

    p <- ggplot(df_pred_ests(), aes(x = df_pred_ests()$time)) +
      geom_point(aes(y = df_pred_ests()$absorbance), shape = 21)+
      geom_path(aes(y = df_pred_ests()$pred), color = 'blue', size = 1) +
      geom_path(aes(y = df_pred_ests()$pred_1), color = 'red', size = 1) +
      geom_path(aes(y = df_pred_ests()$pred_2), color = 'red', linetype = 2, size = 1) +
      facet_wrap(~well, scales = 'free')+
      labs(x = 'Time (Hrs)', y = 'Absorbance') + theme_bw()

  }
  
  if(input$model == 'spline'){
      
      df_ests_spline <- df_ests() %>% pivot_wider(id_cols = well, names_from = term, values_from = estimate)
      
      p <- ggplot(data = df_preds(), aes(x=time, y = pred)) + 
        geom_point(aes(y = absorbance), shape = 21) +
        geom_path(color = 'blue', size = 1)  + 
        geom_abline(data = df_ests_spline, aes(slope = Mu, intercept = -Lag * Mu), color = 'red') + 
        geom_hline(data = df_ests_spline, aes(yintercept = A), color = 'red') + 
        facet_wrap(~well, scales = 'free') + 
        labs(x = 'Time (Hrs)', y = 'Absorbance') + theme_bw()

  }
  
  
  print(p)
        
    })

```

### Error Plots

```{r}

renderPlot({
  
  df_errors <- df_models() %>% filter(fit_error_occured == T)
  
  df_error_plot <- left_join(df_errors, selectedData(), by = 'well')
  
  p <- ggplot(data = df_error_plot, aes_string(x=df_error_plot$time, y = df_error_plot$absorbance)) + 
    geom_point(shape = 21) + 
    facet_wrap(~well, scales = 'free') +
    labs(x = 'Time (Hrs)', y = 'Absorbance') + theme_bw()
  
  print(p)
})

```


## Table{data-height='75%'}

### Error Table

```{r}
renderTable({df_models() %>% select(well, fit_error_occured)})
```

# Parameter Estimates

### Plot estimates{data-height='100%'}
```{r}
renderPlot({
        
        p2 <- ggplot(data = df_ests(), aes(x=df_ests()$well, y = df_ests()$estimate)) + geom_point(size = 3)+
            facet_wrap(~term, scales = 'free') +
            labs(x = 'Well', y = 'Parameter Estimate') + theme_bw()
        
        if(df_ests()$est_error_occured == F){
          p2 = p2 + geom_pointrange(aes(ymin = df_ests()$conf.low, ymax = df_ests()$conf.high )) 
        }
        
        print(p2)
        
    })

```

### Table estimates{data-height='25%'}

```{r}
renderTable({df_ests() %>% select(well, term, estimate, est_error_occured)})
```
